---
title: "SMART Question III"
author: "Andrea Piolini"
date: "November 2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# SMART QUESTION III: 
## Does GDP determine winning/losing? Is GDP a good predictor to predict whether an athlete wins a medal?

To make things more interesting and to deepen our analysis, we decided to introduce GDP as a new variable in this project. We would like to see if GDP has an effect on how well an athlete performs in an event - similar to what we did for sex, age, height, and weight in the previous project. Moreover, we want to see whether GDP influences the gender distribution of a countires athletes, and the overall number of participants per country. Finally, we want to test whether GDP is a good predictor for medals won. In this final part of the analysis we will build several models with GDP included as a feature. In summary, this part of the paper will try to answer the following questions: 

* How does GDP determine statistics in the Olympics?

* Is GDP a good predictor for medals won?

In order to answer these questions, we merged Summer_with_NAs – the dataset on the Olympics we have used so far in our analysis – with the GDP dataset retrieved from the World Bank website. Prior to doing that, we merged Summer_with_NAs with the Name of Country (NOC) dataset – which contains the NOC codes for the countries – by the NOC column. Summer_with_NAs has column called Team which contains the names of the country an athlete comes from, but it also contains names of different teams, such as rowing teams, gymnastic teams, and so on, which is an irrelevant information for our analysis. Therefore, we merged Summer_with_NAs with NOC to have a more accurate columns that contain the country names and we deleted the Team column. Finally, we merged the new dataset with the GDP dataset by country and year. 

```{r init}
loadPkg('readxl')
#loading the datasets
gdp <- read_excel("GDP_Super_Cool.xlsm")
athlete_events <- data.frame(read.csv("athlete_events.csv"))
NOC <- data.frame(read.csv("noc_regions.csv"))
```

```{r cleaning_data}
# keeping NAs for analysis of winners and losers
summer_w_NAs = athlete_events[athlete_events$Season=='Summer',]

# differenciate between winners and losers
# THIS IS VERY IMPORTANT PART OF THE CODE
summer_w_NAs <- subset(athlete_events, Season == 'Summer')
summer_w_NAs$Medal <- as.character(summer_w_NAs$Medal)
summer_w_NAs$Medal[is.na(summer_w_NAs$Medal)] <- "Loser"
summer_w_NAs$Medal[summer_w_NAs$Medal == 'Bronze'] <- 'Winner'
summer_w_NAs$Medal[summer_w_NAs$Medal == 'Silver'] <- 'Winner'
summer_w_NAs$Medal[summer_w_NAs$Medal == 'Gold'] <- 'Winner'
```

```{r merging_dataframe_with_NOC}
#merging the NOc dataset with Summer_w_NAs by NOC
SummerNOC = merge(summer_w_NAs, NOC, by=c("NOC"))
str(SummerNOC)
```

```{r missing_values}
#identifying the missing values for the region column
sum(is.na(SummerNOC$region))
regionNA <- SummerNOC[is.na(SummerNOC$region),]
regionNA
```

```{r deleting_NAs}
#decided to delete the NAs in the region column as they are not relevant for the analysis
SummerNOC <- SummerNOC[!is.na(SummerNOC$region), ]
SummerNOC <- SummerNOC[ , -8]
SummerNOC <- SummerNOC[ , -16]
#renaming the region column to Country
colnames(SummerNOC)[which(names(SummerNOC) == "region")] <- "Country"
```

```{r renaming_countries}
#renaming some of the countries in the gdp dataset as they have a different name in the SummerNOC dataset. This will allow us to create a more accurate dataframe when we merge gdp and SummerNOC
gdp$Country[gdp$Country == 'Egypt, Arab Rep.'] <- 'Egypt'
gdp$Country[gdp$Country == 'Bahamas, The'] <- 'Bahamas'
gdp$Country[gdp$Country == 'Iran, Islamic Rep.'] <- 'Iran'
gdp$Country[gdp$Country == "Cote d'Ivoire"] <- 'Ivory Coast'
gdp$Country[gdp$Country == 'Kyrgyz Republic'] <- 'Kyrgyzstan'
gdp$Country[gdp$Country == 'North Macedonia'] <- 'Macedonia'
gdp$Country[gdp$Country == "Korea, Dem. People's Rep."] <- 'North Korea'
gdp$Country[gdp$Country == 'Russian Federation'] <- 'Russia'
gdp$Country[gdp$Country == 'Slovak Republic'] <- 'Slovakia'
gdp$Country[gdp$Country == 'Korea, Rep.'] <- 'South Korea'
gdp$Country[gdp$Country == 'Syrian Arab Republic'] <- 'Syria'
gdp$Country[gdp$Country == 'Trinidad and Tobago'] <- 'Trinidad'
gdp$Country[gdp$Country == 'United Kingdom'] <- 'UK'
gdp$Country[gdp$Country == 'United States'] <- 'USA'
gdp$Country[gdp$Country == 'Venezuela, RB'] <- 'Venezuela'
gdp$Country[gdp$Country == 'Virgin Islands (U.S.)'] <- 'Virgin Islands, US'
```

```{r new_dataset}
#creating a new data frame SummerGDP by merging SummerNOC and gdp by Year and Country
Summer_GDP = merge(SummerNOC, gdp, by=c("Country", "Year"))
str(Summer_GDP)
```

```{r new_dataset_1}
#scaling GDP, converting Medal to factor, and removing NAs
Summer_GDP$GDP <- scale(Summer_GDP$GDP)
Summer_GDP$Medal <- as.factor(Summer_GDP$Medal)
sum(is.na(Summer_GDP))
Summer_GDP_nona <- na.omit(Summer_GDP)
```

```{r new_dataset_2}
#displaying the number of rows of the new data frame withou NA
nrow(Summer_GDP_nona)
```

It is relevant to mention that some data cleaning was necessary to make the new data frame easier to work on. Additionally, we scaled the GDP variable using the in-built R function scale(), which scaled it on a scale from -1 to 6. Roughly speaking, each GDP unit is worth 300 billion dollars. We also deleted all the rows with NAs to work on a more complete data frame. The following is the structure of Summer_GDP_nona, the data frame we will be using for most of the analysis in this part:  

```{r Summer_GDP_nona_Structure, include = TRUE, echo = FALSE}
#displaying the number of rows of the new data frame withou NA
str(Summer_GDP_nona)
```
### Does GDP determine winning/losing?

The first questions we attempt to answer in this part is if GDP influences whether an athlete earns a medal and if GDP can be a good predictor for that. To see if GDP has an effect on whether an athlete wins a medal we firstly display the data in a boxplot. 


```{r graph1, include=TRUE, echo=FALSE}
#creating a bix plot that displays Winner/Losers and GDP
plot(GDP ~ Medal, data = Summer_GDP_nona, main = "GDP for Losers and Winners", xlab = "Loser/Winner", ylab = "GDP", col = c("darkblue", "antiquewhite"))
```

Next, we run an ANOVA test, which is the best test in this case because we need to compare the means of a numerical variable and a categorical variable.

```{r Anova1}
#Performing ANOVA test on GDP and Medal
GDP_aov <- aov(GDP ~ Medal, data = Summer_GDP_nona)
names(GDP_aov)
summary(GDP_aov)
```

The null hypothesis of this test is the following: the people who won a medal and people who did not win a medal have the same mean for GDP. 

From the ANOVA test we got a really small p-value which is way smaller than our confidence level 0.05 and therefore we reject the null hypothesis and conclude that GDP has an effect on whether an athlete earns a medal. As we can see visually from the boxplot, athletes who won a medal tend to come from countries with higher GDP. 

### Is GDP a good predictor to predict whether an athlete wins a medal?

#### Logistic Regression

Then, we tested GDP as a predictor by building a Generalized Linear Model with scaled GDP as a predictor and Winner/Loser as the Y variable. In this case, the Generalized Linear Model is the best model to use because we are trying to predict Winner/Loser, which is a dichotomous variable. 

```{r logistic_regression, include = TRUE, echo = FALSE}
#performing Logistic regression on GDP and Medal
GDPLogit <- glm(Medal ~ GDP, data = Summer_GDP_nona, family = "binomial")
summary(GDPLogit)
```

From this model we can see that for every unit gain in scaled GDP, the ln(odds-ratio) of winning a medal increases by 0.3. All the p-values that we got in this model are very small and that means that the values we found are statistically significant.

To evaluate whether our model is good, we are going to perform the following tests: 

* Receiver-Operator-Characteristic curve and Area-Under-Curve:
```{r ROC}
loadPkg("pROC")
prob = predict(GDPLogit, type = c("response"))
Summer_GDP_nona$prob=prob
g <- roc(Medal~prob, data= Summer_GDP_nona)
auc(g)
```

```{r AUC, include = TRUE, echo= FALSE}
plot(g)
```

The area-under-curve for this model is `r auc(g)`, which is less than 0.8. Althought `r auc(g)` is not much smaller than 0.8, this means that this model is not a good fit.

```{r McFadden}
loadPkg("pscl")
GDPLogitpr2 = pR2(GDPLogit)
GDPLogitpr2
```

* McFadden: with the McFadden value of `r GDPLogitpr2['McFadden']` only 2% of the variation in y is explained by the explanatory variables in the model, which is not good. 

Since the first model we built is not a good model overall, we built a second Generalized Linear Model including more variables to attempt to create a better model. We included GDP, Weight, Age, and Sex, and we left Height out as it is strongly correlated with Weight. 

```{r logistic_regression2, include = TRUE, echo= FALSE}
#running logistic regression with more variables to see whether it improves the model
GDPLogit1 <- glm(Medal ~ GDP+Weight+ Age + Sex, data = Summer_GDP_nona, family = "binomial")
summary(GDPLogit1)
```

From this model we can see that almost all the variables have positive values, apart from Sex, which is negative. According to the model, if an athlete is a male the ln(odds-ratio) of his chances of winning a medal decrease by 0.5, which seems a little off, as male athletes won more medals than female athletes. In addition, we got significant p-values for all the variables but Age, whose p-value is 0.4, and therefore it is not statistically significant. 

To evaluate whether our model is good, we are going to perform the following tests: 

* Receiver-Operator-Characteristic curve and Area-Under-Curve:
```{r ROC1}
prob1 = predict(GDPLogit1, type = c("response"))
Summer_GDP_nona$prob1=prob1
h <- roc(Medal~prob1, data= Summer_GDP_nona)
auc(h)
```

```{r AUC1, include = TRUE, echo= FALSE}
plot(h)
```

The area-under-curve for this model is `r auc(h)`, which is less than 0.8. Althought `r auc(h)` is not much smaller than 0.8, this means that this model is not a good fit.

```{r McFadden1}
loadPkg("pscl")
GDPLogitpr2_1 = pR2(GDPLogit1)
GDPLogitpr2_1
```

* McFadden: In this case, we got a McFadden value of `r GDPLogitpr2_1['McFadden']`, whihc means that only 3% of the variation in y is explained by the explanatory variables in the model, which is not good.

```{r Q8c}
AIC(GDPLogit)
AIC(GDPLogit1)
```

* We are aslo going to run the Aikake information criterion to compare the two models and see which one of the two is better. We got `r AIC(GDPLogit)` for the first model and `r AIC(GDPLogit1)` for the second model.That means that the second model is better as its value is lower. However, overall our second model is not a good model either. To improve it, we would probably need to add additional variables to the model.

#### k-Nearest Neighbors (KNN)

As logistic regression did not prove to be a good model to predict Winner/Loser, we decided to build another model using the k-Nearest Neighbors (KNN) algorithm. We ran three KNN models using the following three different values: 3,5, and 7. To build them, we used Age, Height, Weight, and GDP as predictors and Winner/Loser as the Y variable.

```{r starting_KNN}
loadPkg("FNN")
#scaling the data
scaledoly1 <- as.data.frame(scale(Summer_GDP_nona[c(7:9, 16)], center = TRUE, scale = TRUE))
```
 
```{r setting_sample}
set.seed(1)
oly_sample1 <- sample(2, nrow(scaledoly1), replace=TRUE, prob=c(0.67, 0.33))
oly_sample1
scaledoly1
```

```{r creating_training_and_test}
oly_training1 <- scaledoly1[oly_sample1==1, 1:3]
oly_test1 <- scaledoly1[oly_sample1==2, 1:3]
```

```{r creating_YValues}
#Now we need to create our 'Y' variables or labels need to input into the KNN function
oly.trainLabels1 <- Summer_GDP_nona[oly_sample1==1, 15]
oly.testLabels1 <- Summer_GDP_nona[oly_sample1==2, 15]
```

```{r K3}
#So now we will deploy our model 
oly_pred1 <- knn(train = oly_training1, test = oly_test1, cl=oly.trainLabels1, k=3)
oly_pred1
```

The cross table for the first model, with k=3, is the following:

```{r loading_gmodels}
loadPkg("gmodels")
```

```{r K3_crosstable, include= TRUE, echo = FALSE}
OLYPREDCross1 <- CrossTable(oly.testLabels1, oly_pred1, prop.chisq = FALSE)
```

```{r K5}
oly_pred2 <- knn(train = oly_training1, test = oly_test1, cl=oly.trainLabels1, k=5)
oly_pred2
```

The cross table for the first model, with k=5, is the following:

```{r K5_crosstable, include = TRUE, echo= FALSE}
OLYPREDCross2 <- CrossTable(oly.testLabels1, oly_pred2, prop.chisq = FALSE)
```

```{r K7}
oly_pred3 <- knn(train = oly_training1, test = oly_test1, cl=oly.trainLabels1, k=7)
oly_pred3
```

The cross table for the first model, with k=7, is the following:

```{r K7_crosstable, include = TRUE, echo= FALSE}
OLYPREDCross3 <- CrossTable(oly.testLabels1, oly_pred3, prop.chisq = FALSE)
```

```{r acc1}
kNN_acc1 =  100 * sum(oly.testLabels1 == oly_pred1)/NROW(oly.testLabels1)
kNN_acc1
```

```{r acc2}
kNN_acc2 =  100 * sum(oly.testLabels1 == oly_pred2)/NROW(oly.testLabels1)
kNN_acc2
```

```{r acc3}
kNN_acc3 =  100 * sum(oly.testLabels1 == oly_pred3)/NROW(oly.testLabels1)
kNN_acc3
```

The first KNN model, with k=3, has an accuracy of `r kNN_acc1`. The second model, with k=5, has an accuracy of `r kNN_acc2`. The third model, with k=7, has an accuracy of `r kNN_acc3`. They are all rather accurate, but the third model is the most accurate of the three. As we can see from the cross tables, all the three models are good at predicting losers and not very good at predicting winners. If we applied it to the real world, it would make sense as just a small part of the athletes wins a medal, whereas the majority does not. 

## Does GDP have an effect on whether and athlete is male or female?

In the second part of this section, we tested whether GDP has an effect on the gender of the athletes. Again, we display the data in a boxplot and then we run an ANOVA test, which is the best test as we are comparing the means of a numerical variable and a categorical variable. 

```{r graph2, include = TRUE, echo= FALSE}
#plotting GDP and Sex in a boxplot
plot(GDP ~ Sex, data = Summer_GDP_nona, main = "GDP for Men and Women", xlab = "Men/Women", ylab = "GDP", col = c("deeppink1", "dodgerblue1"))
```

```{r ANOVA2}
#running ANOVA test for GDP and Sex
GDP_Sex_aov <- aov(GDP ~ Sex, data = Summer_GDP_nona)
names(GDP_Sex_aov)
summary(GDP_Sex_aov)
```

The null hypothesis of this test is the following: the female athletes and male athletes have the same mean for GDP. 

From the ANOVA test we got a really small p-value which is way smaller than our confidence level 0.05 and therefore we reject the null hypothesis and conclude that GDP has an effect on whether female athletes partecipate to the olympics. As we can see visually from the boxplot, the GDP mean for female athletes is higher than the GDP mean for male athletes. Therefore we can conclude that female athletes tend to partecipate more often if they come from a country with a higher GDP. 

## Does GDP have an effect on the number of participants?

Finally, in the last part of this section, we tested whether GDP influences the number of participants to the Olympic Games. To conduct this analysis, we created a new data frame called SummerGDPFreq by creating a frequency table of countries by year and merging it with the GDP dataset. As displayed below, this new data frame no longer contains the name of the athletes along with their features, but it only has the name of the countries, the year, the number of participants for each country each year and GDP.

```{r frequency_table}
#creating a frequency table with the number of participants for each country over the years
table1 <- table(Summer_GDP_nona$Country, Summer_GDP_nona$Year)
table1 <- data.frame(table1) #converting the table into a data frame
#renaming the columns in table1. The Freq column actually indicates the number of participants for each country each year and therefore it was named Participants
colnames(table1)[which(names(table1) == "Var1")] <- "Country"
colnames(table1)[which(names(table1) == "Var2")] <- "Year"
colnames(table1)[which(names(table1) == "Freq")] <- "Participants"
```

```{r another_new_dataframe}
#merging table1 and gdp to create a new data frame SummerGDPFreq 
SummerGDPFreq = merge(table1, gdp, by=c("Country", "Year"))
sum(is.na(SummerGDPFreq$GDP))
SummerGDPFreq <- SummerGDPFreq[!is.na(SummerGDPFreq$GDP), ]
SummerGDPFreq$GDP <- scale(SummerGDPFreq$GDP)
```

```{r displaying_SummerGDPFreq, include= TRUE, echo = FALSE}
str(SummerGDPFreq)
head(SummerGDPFreq)
```

```{r correlation_table}
GDPFreqcor <- cor(SummerGDPFreq[3:4])
GDPFreqcor
```

The following graphs present the correlation among the variables in "SummerGDPFreq" in different ways:

```{r corrplot, include = TRUE, echo= FALSE}
loadPkg("corrplot")
corrplot(GDPFreqcor)
corrplot(GDPFreqcor, method = "number")
```

We the built a linear regression model using scaled GDP as the predictor and Number of Participants as the dependent variable. In this case, the linear model was the best model to use as we are trying to predict a numerical variable which is correlated with its predictor. 

```{r lm1, include = TRUE, echo= FALSE}
model1 <- lm(Participants ~ GDP, data = SummerGDPFreq)
summary(model1)
```

```{r lm1_slope,}
slope <- coef(model1)
slope
```

After we ran our linear regression model, we got an intercept of `r format(coef(model1)[1])` and a slope of `r format(coef(model1)[2])`. This means that for an increase of one point for the scaled GDP, which is worth billions of dollars, we expect an increase of more than 72 participants in total on average. Additionally, we got a really tiny p-value for the slope. Therefore, we reject the null hypotheisis, which is that the intercept and the slope is equal to 0, and we say that our slope is statistically significant. 

Finally, we got an R-squared value of 0.336. This means that our model accounts for approximately 34% of the variance.

```{r lm1_plotted}
plot(model1)
loadPkg("ggplot2")
ggplot(model1,aes(GDP,Participants))+geom_point(aes(GDP,Participants)) + geom_line(aes(Participants), colour="red", size=1)
```